---
title: "Assignment 3: Geospatial Data Sciences and Economic Spatial Models"
author: |
  Sebastian Dong Uk Paik Sohn, Michael Duarte Gonçalves  
  Gal·la Gelpí Buxadé, Maria Victoria Suriel Nuñez
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 1
    number_sections: true
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{xcolor}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, urlcolor=green!50!black}
  - \usepackage{tocloft}
  - \renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}
---

\newpage

# Setup

We load `gdistance` last because it masks several `tidyverse` functions
(`select`, `filter`, `extract`, `union`). To stay safe, we prefix all
ambiguous calls with `dplyr::`.

```{r packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(spData)
library(rnaturalearth)
library(rnaturalearthdata)
library(gdistance)
```

The SPEI dataset lives on the course Dropbox. We only download it once.

```{r download-spei}
dir.create("data", showWarnings = FALSE)

spei_path <- "data/spei01.nc"

if (!file.exists(spei_path)) {
  message("Downloading SPEI data (~340 MB)...")
  download.file(
    url = str_c(
      "https://www.dropbox.com/scl/fi/",
      "kmo2gj0iqvu52rat9ydmt/",
      "spei01.nc?rlkey=1ahg7k4cs1uf1v52s5x6d0sxj",
      "&dl=1"
    ),
    destfile = spei_path,
    mode     = "wb"
  )
}
```

\newpage

# Part 1: Climate Change in the USA

## Preparing the regions

We start from `us_states` (spData), drop Alaska and Hawaii, and dissolve
state boundaries into the four Census regions.

```{r us-regions}
# Keep continental states with their region label
sf_states <- us_states %>%
  dplyr::filter(!NAME %in% c("Alaska", "Hawaii")) %>%
  dplyr::select(NAME, REGION)

# Merge states into region-level polygons
sf_regions <- sf_states %>%
  group_by(REGION) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup()
```

## Loading the SPEI raster

The NetCDF has 1,380 monthly layers (Jan 1901 - Dec 2015). We keep all
months from 1966 to 2015 and compute the yearly average, giving us a
clean 50-year annual panel.

```{r load-spei}
r_spei     <- rast(spei_path)
spei_dates <- time(r_spei)

# Keep all monthly layers from 1966 to 2015
layer_index <- which(
  year(spei_dates) >= 1966 &
    year(spei_dates) <= 2015
)

r_spei_monthly <- r_spei[[layer_index]]

# Compute yearly averages (mean of 12 months per year)
spei_years_all <- year(time(r_spei_monthly))
r_spei_50 <- tapp(
  r_spei_monthly, spei_years_all, fun = mean
)

spei_years <- names(r_spei_50) %>%
  str_extract("\\d{4}") %>%
  as.integer()

names(r_spei_50) <- str_c("spei_", spei_years)
```

## Zonal statistics

We compute the area-weighted mean SPEI within each region for every year.
This gives us a panel dataset, essentially one number per region per year, 
which we then reshape into long (tidy) format for plotting.

```{r zonal-stats, message = FALSE, warning = FALSE}
# Area-weighted mean SPEI per region per year
df_zonal <- exact_extract(
  x   = r_spei_50,
  y   = sf_regions,
  fun = "mean"
)

# Pivot into a tidy panel: one row per region-year
df_panel <- df_zonal %>%
  mutate(REGION = sf_regions$REGION) %>%
  pivot_longer(
    cols      = starts_with("mean.spei_"),
    names_to  = "year",
    values_to = "spei"
  ) %>%
  mutate(
    year = str_extract(year, "\\d{4}") %>%
      as.integer()
  )
```

## Figure 1: SPEI evolution by region

We plot the SPEI for each Census region over the 50-year window.
Positive values indicate wetter-than-normal conditions, while negative values
signal drought. We overlay a LOESS smoother (in blue) to capture the overall
trend across all four regions.

```{r fig-01, fig.cap = "SPEI by region.", fig.width = 6.5, fig.height = 4}
ggplot(
  df_panel,
  aes(x = year, y = spei, color = REGION)
) +
  geom_line(alpha = 0.7, linewidth = 0.6) +
  geom_smooth(
    aes(group = 1),
    method    = "loess",
    color     = "blue",
    linewidth = 1.2,
    se        = TRUE
  ) +
  labs(x = "year", y = "spei", color = "REGION") +
  theme_bw()
```

Several patterns stand out. First, we observe a gentle downward slope in
the LOESS trend, suggesting a modest drying tendency over the half-century.
Second, individual drought episodes, like the mid-1970s, the
late 1980s, appear simultaneously in multiple regions, pointing
to large-scale climate drivers rather than purely local phenomena.
Overall, the figure illustrates that while
year-to-year variability dominates, there is a slow structural shift toward
drier conditions that matters for long-run agricultural planning and water
resource management.

\newpage

# Part 2: Transportation Centrality in Spain

## Geographic setup

We grab Spain's boundary from `spData::world`, then download Natural Earth
populated places and roads at 10m resolution. Column names come in lowercase
from the `.gpkg` format, so we normalize just in case.

```{r spain-boundary}
sf_spain <- world %>%
  dplyr::filter(name_long == "Spain") %>%
  st_transform("EPSG:4326")
```

```{r download-places, message = FALSE, warning = FALSE, results = "hide"}
# Populated places
sf_places_raw <- ne_download(
  scale       = 10,
  type        = "populated_places",
  category    = "cultural",
  destdir     = "data",
  returnclass = "sf"
)

# Ensure lowercase columns (gpkg vs shp inconsistency)
names(sf_places_raw) <- tolower(names(sf_places_raw))

# Top 10 Spanish cities by population
sf_places <- sf_places_raw %>%
  dplyr::filter(iso_a2 == "ES") %>%
  arrange(desc(pop_max)) %>%
  slice_head(n = 10) %>%
  dplyr::select(name, pop_max)
```

```{r download-roads, message = FALSE, warning = FALSE, results = "hide"}
# Road network, cropped to Spain
sf_roads_raw <- ne_download(
  scale       = 10,
  type        = "roads",
  category    = "cultural",
  destdir     = "data",
  returnclass = "sf"
)

sf_roads_spain <- sf_roads_raw %>%
  st_crop(st_bbox(sf_spain)) %>%
  st_intersection(sf_spain)
```

## Friction surface

We rasterize the road network at 0.1 degree resolution. Road cells receive
a value of 1, while all other cells get 1/100, meaning off-road travel is
100 times more costly. This is a deliberately simple model: we are not
claiming it captures actual travel times, but it gives us a
reasonable approximation of how the road network shapes
connectivity.

```{r friction-surface}
# Template raster covering Spain
r_template <- rast() %>% crop(sf_spain)
res(r_template) <- 0.1

# Burn roads into the grid
r_roads <- rasterize(
  vect(sf_roads_spain), r_template
)

# Off-road pixels become 1/100 (costly but not impossible)
vv <- values(r_roads)
vv[is.nan(vv) | is.na(vv)] <- 1 / 100
values(r_roads) <- vv
rm(vv)
```

## Figure 2: Friction surface

```{r fig-02, fig.cap = "Friction surface.", fig.width = 5, fig.height = 5}
plot(r_roads, main = "Friction Surface - Spain")
plot(
  st_geometry(sf_spain),
  add = TRUE, border = "black", lwd = 2
)
plot(
  st_geometry(sf_places),
  add = TRUE, pch = 20, cex = 1.5
)
```

## Transition matrix and bilateral distances

We convert our friction raster into a transition matrix using `gdistance`.
We apply a geographic correction to account for the fact that grid cells
at different latitudes have different physical sizes. Then we compute all
pairwise distances in a single call with `costDistance()`. We do this because
it is much faster than looping `shortestPath()` over every pair individually.

```{r transition-matrix}
# gdistance needs the old raster format
tr_matrix <- transition(
  x                  = raster::raster(r_roads),
  transitionFunction = mean,
  directions         = 8
) %>%
  geoCorrection(type = "c")
```

```{r bilateral-distances}
# All pairwise distances in one call
dist_matrix <- costDistance(
  tr_matrix,
  as(sf_places, "Spatial")
) %>%
  as.matrix()

# Label and convert to km
city_names <- sf_places$name
rownames(dist_matrix) <- city_names
colnames(dist_matrix) <- city_names

dist_matrix_km <- dist_matrix / 1000
round(dist_matrix_km, 0)
```

## Figure 3: Madrid--Vigo shortest path

While `costDistance()` gives us the numbers, we also want to see the actual
route on a map. We use `shortestPath()` for the Madrid-Vigo pair specifically,
which returns a line geometry we can overlay on the road network.

```{r shortest-path}
sf_madrid <- sf_places %>%
  dplyr::filter(name == "Madrid")
sf_vigo <- sf_places %>%
  dplyr::filter(name == "Vigo")

# shortestPath gives us the line geometry for plotting
path_mv <- shortestPath(
  x      = tr_matrix,
  origin = st_coordinates(sf_madrid),
  goal   = st_coordinates(sf_vigo),
  output = "SpatialLines"
)

# Convert to sf and assign CRS (shortestPath drops it)
sf_path_mv <- path_mv %>%
  st_as_sf() %>%
  st_set_crs(4326)
```

```{r fig-03, fig.cap = "Madrid--Vigo path.", fig.width = 5.5, fig.height = 5.5}
ggplot() +
  geom_sf(
    data  = sf_spain,
    fill  = "grey90",
    color = "black"
  ) +
  geom_sf(
    data      = sf_roads_spain,
    color     = "grey40",
    linewidth = 0.3
  ) +
  geom_sf(
    data      = sf_path_mv,
    color     = "red",
    linewidth = 1
  ) +
  geom_sf(
    data  = sf_places,
    size  = 2,
    color = "black"
  ) +
  geom_sf(
    data  = sf_madrid,
    size  = 4,
    color = "red"
  ) +
  geom_sf(
    data  = sf_vigo,
    size  = 4,
    color = "blue"
  ) +
  geom_sf_label(
    data    = sf_madrid,
    aes(label = name),
    nudge_y = 0.3
  ) +
  geom_sf_label(
    data    = sf_vigo,
    aes(label = name),
    nudge_y = 0.3
  ) +
  labs(
    title = "Road Network & Shortest Path: Madrid - Vigo"
  ) +
  theme_minimal()
```

```{r path-length}
sf_path_mv %>%
  st_length() %>%
  `/`(1000) %>%
  round(0) %>%
  str_c(
    "Madrid-Vigo network distance: ", ., " km"
  ) %>%
  cat("\n")
```

## Comparing isolation

We extract each city's row from the distance matrix. We then
stack the two sets of distances into a
tidy tibble so we can compare them side by side.

```{r isolation-data}
idx_madrid <- which(city_names == "Madrid")
idx_vigo   <- which(city_names == "Vigo")

df_distances <- bind_rows(
  tibble(
    origin   = "Madrid",
    dest     = city_names[-idx_madrid],
    distance = dist_matrix_km[idx_madrid, -idx_madrid]
  ),
  tibble(
    origin   = "Vigo",
    dest     = city_names[-idx_vigo],
    distance = dist_matrix_km[idx_vigo, -idx_vigo]
  )
)
```

## Figure 4: Distance distributions

We now compare how far each city is from every other top-10 city. The density
plot tells us, at a glance, whether a city sits at the centre of the network
or out on the periphery. If a city's curve is concentrated on the left, most
other cities are nearby -- that city is well connected. If the curve shifts
to the right, the city is structurally more isolated.

```{r fig-04, fig.cap = "Madrid vs Vigo.", fig.width = 6, fig.height = 4}
ggplot(
  df_distances,
  aes(x = distance, fill = origin)
) +
  geom_density(alpha = 0.5) +
  xlim(0, 1500) +
  scale_fill_manual(
    values = c(
      "Madrid" = "orangered",
      "Vigo"   = "forestgreen"
    )
  ) +
  labs(
    title = "Transportation Centrality: Madrid vs. Vigo",
    x     = "Distance (km)",
    y     = "Density",
    fill  = "Origin"
  ) +
  theme_minimal()
```

The result is striking. We see that Madrid's distribution peaks at
shorter distances and is more tightly concentrated, reflecting its position
at the center of Spain's radial highway system. Most major cities are reachable
within a relatively narrow band. Vigo's distribution, by contrast,
is shifted substantially to the right, every other top-10 city is farther
away, and the spread is wider.

We can interpret the gap between the two curves as a measure
of differential market access. Firms in Madrid can
reach a larger consumer base at lower transport cost than firms in Vigo.
This structural advantage helps explain Madrid's dominance as a logistics
and services hub, while Vigo, despite being an important Atlantic port, 
remains more regionally oriented.
